---
title: "Week 3"
output:  Data cleaning
Author: Faria Hossain and Asim Gul
---



##Data cleaning--part 1
I downloaded the dataset from the Berkeley Statistics website and then imported it into RStudio manually using a button in the top right window of the program called “Import Dataset”.

The second way is by downloading data directly into whatever software you’re using for your analysis, in this case RStudio. (Another very popular platform that we won’t be using in this class is the web-based Jupyter Notebook, which you will learn much more about if you start learning Python for data analysis.)
```{r}
url <- 'https://www.stat.berkeley.edu/users/statlabs/data/video.data'

video <- read.table(url)

head(video)
```
```{r}
names(video)
```
```{r}
colnames(video)
```

```{r}
rownames(video)
```

```{r}
summary(video)
```
 
By looking again at the documentation for read.table(), we find that there is a second optional argument for this function called header, which is “a logical variable indicating whether the file contains the names of the variables as the first line”. Its default value is FALSE, and since we didn’t tell R otherwise, it generated new variable names for us, hence these weird numbered variable names that start with V. (“V” probably stands for “vague”.) To fix this, all we need to do is add header = TRUE to our read.table() call.

```{r}
video <- read.table(url, header = TRUE)
```
```{r}
dim(video)
```


```{r}
head(video)
```

```{r}
tail(video)
```
R has more functions for summarizing data that you may find more useful depending on the situation. One is names(), which simply lists the names of all of the variables in a data frame. This is useful if you’re having trouble remembering the exact name of a variable.

```{r}
names(video)
```

Another is summary(), which numerically summarizes all of the variables passed to it.

```{r}
summary(video)
```


## 3.5 Exploratory Data Analysis

The first thing we need to do with our data is have a look at its structure. The str() function is great for this. Lets see what it does!
```{r}
str(video)
```
The output above tells us a lot about our data. Thanks to str() we now know that our data consists of 91 observations of 15 variables, and that the variables are a combination of factors, numerics and integers.


To clean up this dataset, first we need to review using the codebook what each of the variables in it represent. After doing this it is obvious that time, age and work are appropriately expressed as numerics or integers, so we do not need to do modify these variables in any way.

It follows that we need to somehow change all of the other variables. Since we now know that the values of these other variables in their current form correspond to verbal responses to the questions that correspond to these variables, we should change the values of these variables so that they represent the responses themselves and not numbers which represent these responses.

In order to do this, we need to transform these variables into different data types, in this case factors. We can do this using the factor() function. After turning these variables into factors, we need to change the levels of each of these variables so that each level corresponds to a verbal statement instead of just a number. We will start by working with like.

```{r}
video$like <- factor(video$like)
```

levels() provides access to the levels attribute of a variable. The first form returns the value of the levels of its argument and the second sets the attribute.

```{r}
levels(video$like)
```

```{r}
levels(video$like) <- c('Never played', 
                        'Very much', 
                        'Somewhat', 
                        'Not really', 
                        'Not at all',
                        'No response')
levels(video$like)
```
To change the levels of this factor so that they contain levels that are verbal statements and not numbers that represent verbal statements, first we must transform video$like into a factor variable using factor().

According to the codebook, the values of like have the following meanings: “1=Never played, 2=Very much, 3=Somewhat, 4=Not really, 5=Not at all”. These are the levels for like that we want to use in our analysis of this data. But we see from the printout of levels(video$like) that this variable actually has six levels, the last being 99. Recall from the last tutorial that 99 represents No response. We need to include this level too, just like we did in the last tutorial.

Next, we need to overwrite the current levels of this new factor variable, which are still 1, 2, 3, 4, 5 and 99. This information is contained in a vector, so we need to overwrite it using a new vector which contains the same number of elements as the original vector. This detail is very important. If the vector we are using for overwriting contains a different number of elements than the original vector, R will throw an error. Always inspect the levels of a factor variable that you’re modifying before modifying it to avoid simple and frustrating mistakes like this.

For the remaining variables to be modified, the cleaning process is identical, so I will include one more example below but exclude the other cases because it would be too repetitive.

```{r}
video$where <- factor(video$where)
levels(video$where)
```

```{r}
levels(video$where) <- c('Arcade', 
                         'Console', 
                         'Personal computer', 
                         'Computer and console', 
                         'Arcade and home(console or computer)', 
                         'Arcade and home (both console and computer)',
                         'No response')
levels(video$where)
```

After completing the cleaning process for this dataset, we see a familiar sight.

```{r}
str(video)
```




