---
title: "R Notebook"
output: html_notebook
---

### Covariance in R
```{r}
t=seq(0,10,0.5)
y1=sin(t)
y2=cos(t-0.6*pi)
y3=cos(t+0.5*pi)
plot(t,y1,type="o", lty=2, xlab="time", ylab="Sine/Cos wave",col='blue')
points(t, y2, type="o", lty=1, col = 'red')
```
```{r}
cov(y1, y2)  
cor(y1, y2) 
```
```{r}
plot(t,y1,type="o", lty=2, xlab="time", ylab="Sine/Cos wave",col='blue')
points(t, y3, type="o", lty=1, col = 'red')
```
```{r}
cov(y1, y3)  
cor(y1, y3) 
```
## covariance and correlation between X,Y
```{r}
X <- rnorm(100,mean=10,sd=2)
plot(density(X))
```
creates a r.v. from a normal distribution, 100 observations, with mean =10 and s.d.=2
```{r}
Y <- 2*X+rnorm(100,mean=4,sd=1)
plot(density(Y))
```
creates a second r.v. which is twice the first one plus some noise
```{r}
plot(X,Y)
```

```{r}
v_X=var(X)
v_Y=var(Y)
v_XY=var(X+Y)
print(v_X)
print(v_Y)
print(v_XY)
```


```{r}
cov_XY=0.5*(v_XY-v_X-v_Y)
print(cov_XY)
```


```{r}
cov(X,Y)
#gives covariance between X,Y
cor(X,Y)
#gives correlation between X,Y
```

### Negative covariance in R
```{r}
X <- rnorm(100,mean=10,sd=2)
plot(density(X))
# creates a r.v. from a normal distribution, 100 observations, with mean =10 and s.d.=2
```

```{r}
Y <- -2*X+rnorm(100,mean=4,sd=1)
plot(density(Y))
# creates a second r.v. which is twice the first one plus some noise
```
```{r}
plot(X,Y)
```

```{r}
cov(X,Y)
#gives covariance between X,Y
cor(X,Y)
#gives correlation between X,Y
```
## Independent X and Y
```{r}
X <- rnorm(100,mean=10,sd=2)
plot(density(X))
# creates a r.v. from a normal distribition, 100 observations, with mean =10 and s.d.=2
Y <- rnorm(100,mean=4,sd=1)
plot(density(Y))
# creates a second r.v.
plot(X,Y)
```

```{r}
cov(X,Y)
#gives covariance between X,Y
cor(X,Y)
#gives covariance between X,Y
```

### Combinations and Permutations

```{r}
install.packages("combinat")                   # Install combinat package
library("combinat")                            # Load combinat package
```

combn(x, m, fun=NULL, simplify=TRUE, …)

Parameters : 

x – vector source for combinations
m – number of elements to be taken
fun – function to be applied to each combination (may be null)
simplify – logical, if FALSE, returns a list, otherwise returns vector or array

```{r}
permut <- permn(3)                             # Create list of permutations
permut                                         # Print list of permutations
```
```{r}
permut_count <- length(permn(3))               # Count permutations
permut_count                                   # Print count of permutations
```
```{r}
combi <- combn(3, 2)                 # Create matrix of combinations
combi                                          # Print matrix of combinations
```

```{r}
combi_count <- ncol(combn(3, 2))     # Count combinations
combi_count                          # Print count of combinations
```

### The binomial distribution in R
R has several built-in functions for the binomial distribution. They’re listed in a table below along with brief descriptions of what each one does.

X~B(n,p)

##1. rbinom(n, size, prob)	

Generates numbers which follow a binomial distribution with the given parameters

Let’s try these functions out to see how they really work.

We’ll start with rbinom(), a function which randomly generates numbers which follow a binomial distribution with given parameters. For our first test of it, we’ll generate one observation (n = 1) of a sample of size 100 (size = 100) and a probability of success of 0.3 (prob = 0.3).

```{r}
set.seed(10)
rbinom(3, 100, 0.3)
```

```{r}
y2 <- rbinom(8, size = 100, prob = 1 / 6) 
print(y2)
```

##2. dbinom(x, size, prob)

P(X = x), the probability that X = x


Since only one out of five possible answers is correct, the probability of answering a question correctly by random is 1/5=0.2. We can find the probability of having exactly 4 correct answers by random attempts as follows.
```{r}
dbinom(4, size=12, prob=0.2)
```

```{r}
dbinom(4, size=8, prob=0.5)
```

```{r}
# Create a sample of 50 numbers which are incremented by 1.
x <- seq(0,50,by = 1)
# Create the binomial distribution.
#y <- dbinom(x,size=50,prob=0.5)
y <- dbinom(x,50,0.5)
# Plot the graph for this sample.
plot(x,y)
```

```{r}
dbinom(3, size = 13, prob = 1 / 6) 
probabilities <- dbinom(x = c(0:10), size = 13, prob = 1 / 6) 
data.frame(probabilities)
#plot(0:10, probabilities, type = "l") 
plot(0:0.1:10, probabilities, type = "l") 
```

##3. pbinom(q, size, prob, lower.tail = TRUE)

P(X =< q), the probability that X takes a value less than or equal to q


Alternatively, we can use the cumulative probability function for binomial distribution pbinom.

```{r}
pbinom(4, size=8, prob=0.5) 
```
```{r}
x <- pbinom(26,51,0.5)
print(x)
```
```{r}
pbinom(3, size = 13, prob = 1 / 6) 
plot(0:10, pbinom(0:10, size = 10, prob = 1 / 6), type = "l")
```
 
###The geometric Distribution in R
A sports marketer randomly selects persons on the street until he encounters someone who attended a game last season. What is the probability the market encounters x = 10 people who did not attend a game before the first success when p = 0.20 of the population attended a game?
```{r}
p = 0.20
n = 10
# exact
dgeom(x = n, prob = p)
```
```{r}
rgeom(n = 100, prob = p)
```



```{r}
# simulated
mean(rgeom(n = 100, prob = p) == 10)
```
```{r}
library(dplyr)
library(ggplot2)
```

Example
From the prior example, what is the probability the marketer fails to find someone who attended a game in x <= 5 trials before finding someone who attended a game on the next trial when the population probability is p = 0.20?
```{r}
p = 0.20
n = 5
# exact
pgeom(q = n, prob = p, lower.tail = TRUE)
```
```{r}
# simulated
mean(rgeom(n = 10000, prob = p) <= n)
```

```{r}
library(dplyr)
library(ggplot2)

data.frame(x = 0:10, 
           pmf = dgeom(x = 0:10, prob = p),
           cdf = pgeom(q = 0:10, prob = p, lower.tail = TRUE)) %>%
  mutate(Failures = ifelse(x <= n, n, "other")) %>%
ggplot(aes(x = factor(x), y = cdf, fill = Failures)) +
  geom_col() +
  geom_text(
    aes(label = round(cdf,2), y = cdf + 0.01),
    position = position_dodge(0.9),
    size = 3,
    vjust = 0
  ) +
  labs(title = "Cumulative Probability of X = 5 Failures.",
       subtitle = "Geometric(.3)",
       x = "Failures prior to first success (x)",
       y = "probability") 
```

