---
title: "Week 7"
output: Hypergeometric distribution/Possion distribution 
Author: Faria Hossain
---


### Covariance in R
```{r}
t=seq(0,10,0.2)
y1=sin(t)
y2=cos(t-0.55*pi)
y3=cos(t+0.5*pi)
y4=0.5*sin(t)
plot(t,y1,type="o", lty=2, xlab="time", ylab="Sine/Cos wave",col='blue')
points(t, y2, type="o", lty=1, col = 'red')
points(t, y3, type="o", lty=3, col = 'black')
points(t, y4, type="o", lty=4, col = 'green')
```
```{r}
plot(y1,y2)
```



```{r}
cc12<-c(cov(y1, y2),cor(y1, y2))
cc13<-c(cov(y1, y3),cor(y1, y3))
cc14<-c(cov(y1, y4),cor(y1, y4))
cc12
cc13
cc14
```


```{r}
plot(t,y1,type="o", lty=2, xlab="time", ylab="Sine/Cos wave",col='blue')
points(t, y3, type="o", lty=1, col = 'red')
```

```{r}
data()
```

```{r}
head(sleep)
```

```{r}
cov(y1, y3)  
cor(y1, y3) 
```
## covariance and correlation between X,Y
```{r}
X <- rnorm(100,mean=10,sd=2)
X
plot(density(X))
```

## Hypergeometric distribution in R

# Example 1:  dhyper(x, m, n, k) 
It is defined as Hypergeometric Density Distribution used in order to get the density value.

x, q	
vector of quantiles representing the number of white balls drawn without replacement from an urn which contains both black and white balls.

m	--> r
the number of white balls in the urn.

n	--> N-r
the number of black balls in the urn.

k	-->n
the number of balls drawn from the urn, hence must be in 0,1,…, m+n.

##page 10, example 1
```{r}
x = 2
r = 6
n = 10
k = 5
dhyper(0, r, n, k )
```
```{r}
k * r / (r + n)
```

```{r}
# Specify x-values for dhyper function 
x<- seq(0, 5, by = 1) 
r = 6 ## r=6
n = 10 ## N-r=16-6=10
k = 5 ## n=5
y<- dhyper(x, r, n, k )
plot(x,y,xlab = 'Number of underweight milk containers found by inspector',ylab='Probability',type='h') 
text(x, y, round(y,digits=3), cex=0.7, pos=4, col="red")
```


##page 11
```{r}
m = 7
n = 20-7
k = 5
dhyper(x = 1, m, n, k )
```

##page 12
```{r}
x = 4
m = 6
n = 14
k = 5
dhyper(x, m, n, k )
```

# Example 2:  phyper(x, m, n, k)

Hypergeometric Cumulative Distribution Function used estimating the number of faults initially resident in a program at the beginning of the test or debugging process based on the hypergeometric distribution and calculate each value in x using the corresponding values.

##page 10
```{r}
# Specify x-values for phyper function 
x<- seq(0, 5, by = 1) 
m = 6
n = 10
k = 5    
# Apply phyper function 
y_phyper <- phyper(x, m, n , k )   
# Plot phyper values 
plot(x,y_phyper,type = 'l')   
```

# Example 3:  rhyper(N, m, n, k)

It generally refers to generating random numbers function by specifying a seed and sample size.

```{r}
# Set seed for reproducibility 
# Specify sample size 
set.seed(400)                                  
N <- 10000                                       
  
# Draw N hypergeometrically distributed values 
y_rhyper <- rhyper(N, m = 50, n = 20, k = 30)  
y_rhyper          
  
# Plot of randomly drawn hyper density 
hist(y_rhyper,                                           
     breaks = 50, 
     main = "") 
```

## The Poisson distribution in R

R has several built-in functions for the Poisson distribution. They’re listed in a table below along with brief descriptions of what each one does.

Poisson function	What it does


#Example 1:   rpois(n, lambda)	
Returns n randomly generated numbers that follow a Poisson distributionn with an average number of lambda successes
We will begin our demom with rpois(). First we’ll use it to make one (n = 1) randomly generated observation of a random variable that follows the Poisson distribution and has an average number of 10 successes (lambda = 10) per period.


```{r}
rpois(10, 10)
```

# Example 2:  dpois(x, lambda)	
P(X = x), the probability that there will be x successes per period for an event with an average number of lambda successes

##page 18
```{r}
dpois(x = 4, lambda = 1.8) 
```

##page 19
```{r}
1-(dpois(x = 0, lambda = 1.8) +
  dpois(x = 1, lambda = 1.8))
```
## page 22
```{r}
events <- 0:20
density1 <- dpois(x = events, lambda = 3)
density2 <- dpois(x = events, lambda = 5)
density3 <- dpois(x = events, lambda = 10)
prob1 <- ppois(q = events, lambda = 3, lower.tail = TRUE)
par(mfrow=c(1,3)) 
plot(events,density1,type='h',main = "Po(3)",ylab = "P(X)",xlab = "x",ylim=c(0,0.25))
plot(events,density2,type='h',main = "Po(5)",ylab = "P(X)",xlab = "x",ylim=c(0,0.25))
plot(events,density3,type='h',main = "Po(10)",ylab = "P(X)",xlab = "x",ylim=c(0,0.25))
```




## Example 3: ppois(x, lambda, lower.tail = TRUE)	

P(X <= x), the cumulative probability that there will be x or fewer successes per period for an event with an average number of lambda successes. Returns P(X > x), the cumulative probability that there will be more than x successes per period for the same variable when lower.tail = FALSE.

What is the probability of making 2 to 4 sales in a week if the average sales rate is 3 per week?
```{r}
# Using cumulative probability
ppois(q = 4, lambda = 3, lower.tail = TRUE) - 
ppois(q = 2, lambda = 3, lower.tail = TRUE)
```
```{r}
x <- c(0,1,2,5,8,10,15,20)
ppois(x,6)
```

## cdf of page 22
```{r}
events <- 0:20
density1 <- dpois(x = events, lambda = 3)
density2 <- dpois(x = events, lambda = 5)
density3 <- dpois(x = events, lambda = 10)
prob1 <- ppois(q = events, lambda = 3, lower.tail = TRUE)
prob2 <- ppois(q = events, lambda = 5, lower.tail = TRUE)
prob3 <- ppois(q = events, lambda = 10, lower.tail = TRUE)
par(mfrow=c(1,3)) 
plot(events,prob1,type='l',main = "Po(3)",ylab = "P(X)",xlab = "x")
plot(events,prob2,type='l',main = "Po(5)",ylab = "P(X)",xlab = "x")
plot(events,prob3,type='l',main = "Po(10)",ylab = "P(X)",xlab = "x")
```
```{r}
1-ppois(q = 3, lambda = 2)
```

```{r}
##Example of disease incidence
events <- 0:10
density <- dpois(x = events, lambda = 2)
prob1 <- ppois(q = events, lambda = 2, lower.tail = TRUE)
par(mfrow=c(1,2)) ## Generate two sub plot.
plot(events,density,type='h',main = "PMF:Po(2)",ylab = "P(X)",xlab = "x")
plot(events,prob1,type='b',main = "CDF:Po(2)",ylab = "P(X)",xlab = "x")
```


```{r}
# expected number of sales = lambda = 3
# variance = lambda = 3
#install.packages('dplyr')
library(ggplot2)
library(dplyr)
options(scipen = 999, digits = 2) # sig digits
events <- 0:10
density <- dpois(x = events, lambda = 3)
prob <- ppois(q = events, lambda = 3, lower.tail = TRUE)
df <- data.frame(events, density, prob)
ggplot(df, aes(x = factor(events), y = density)) +
  geom_col() +
  geom_text(
    aes(label = round(density,2), y = density + 0.01),
    position = position_dodge(0.9),
    size = 3,
    vjust = 0
  ) +
  labs(title = "PMF and CDF of Poisson Distribution",
       subtitle = "P(3).",
       x = "Events (x)",
       y = "Density") +
  geom_line(data = df, aes(x = events, y = prob))
```


## Random Sample

```{r}
weight <- rnorm(10000,78,sd=5) #simulate 10000 measurements of weight with mean 78kg
hist(weight) # a histogram
```

```{r}
Sa <- NULL # take some random samples
for(i in 1:40){ # forty of them
jj <- sample(weight,50) # of size 50
Sa <- rbind(Sa,jj)
} # and store the outcome
xbar <- apply(Sa,1,mean) # find the mean in each sample
varbar<- apply(Sa,1,var) # and the variance
plot(xbar,ylim=c(60,90)) #plot the sample mean
sampleMean <- mean(xbar) #find the means of all samples
```

```{r}
mean(xbar) #find the means of all samples
var(xbar)
```
## Makcov's abd Chebyshev's

## page 28
```{r}
mu=1.3
var=0.25
x <- seq(0, 5, by = .1)
y <- dnorm(x, mean = mu, sd = var)
plot(x,y)
```

pnorm(q, mean = 0, sd = 1, lower.tail = TRUE)	Calculates P(X <= x) for a given mean and standard deviation. Default mean is 0 and default standard deviation is 1. Calculates P(X > x) when lower.tail = FALSE.
```{r}
mu=1.3
var=0.25
x=3
p1<-pnorm(x, mean = mu, sd = var,lower.tail = FALSE) 
c1<-mu/x
p1
c1
p1 < c1
```

```{r}
mu=1.3
var=0.25
x=2
2*pnorm(x*var, mean = mu, sd = var,lower.tail = TRUE) 
1/x^2
```


### Law of Large Numbers Simulations

##example 1

```{r}
a<-pbinom(50, 100, 0.5)
round(a,10)
```
##example 2


```{r}
mu <- 34.5
sigma <- 14.15
```

Next, we will create a vector named mean_vector to hold our sample means:
```{r}
mean_seq <- rep(0, 30)
```
The rep function, as used above, creates a vector of length 30 that initially contains all zero values.
Every time we calculate a sample mean, we will store it in this vector.
Now you will proceed to create 30 random samples from our normal distribution, and these samples
will increase in size from n=1 to n=30. For the first sample, type:

```{r}
sample <- rnorm(1, mean=mu, sd=sigma)
mean_seq[1] <- mean(sample)
mean_seq[1]
```

The first line uses the rnorm function to draw a sample of size 1 from our normal distribution with the mu and sigma parameters we specified earlier. The second line uses the mean function to calculate the arithmetic mean of the sample and then stores it in the first index of the mean_vector variable.

To fill in the rest of the vector, we will increase the sample size by one each time. 

To save some typing, however, we will combine the process of drawing the random sample, calculating the sample mean, and storing the value in mean_vector into a single step. Thus, for a sample size of 2, type the following:
```{r}
mean_seq[2] <- mean(rnorm(2, mean=mu, sd=sigma))
```

The rnorm function will draw a sample of size 2 from our hypothetical distribution, the mean function will then calculate the sample mean, and then the mean will get stored in the second index  of the mean_vector variable. 

Now for the tedious part: you will need to repeat this process for all of the remaining sample sizes. Use the Up-Arrow key and then change the index position and the sample size, as illustrated below:

```{r}
x=5000
mean_seq <- rep(0, x)

for(a in 1:x){ # forty of them
mean_seq[a] <- mean(rnorm(a, mean=mu, sd=sigma))
}
```
Finally, we will create a scatterplot of the results using the plot function:

```{r}
plot(seq(1,x), mean_seq)
abline(h=mu, col="red")
```