---
title: "R Notebook"
output: Calculate Mean, Median, Mode, Variance, Standard Deviation

---

```{r}
3924600%%6
```

## Example page 12
```{r}
f=function(x) 2/((x+1)^3)
integrate(f,0,Inf) 
```
```{r}
E_f=function(x) x*2/((x+1)^3)
integrate(E_f,0,Inf) 
```

## Example page 13
```{r}
f2=function(x) 0.128*exp(-x/100)
integrate(f2,120,150) 
```
```{r}
E_f2=function(x) x*0.128*exp(-x/100)
integrate(E_f2,120,150) 
```

## Mean and Variance

```{r}
x_1 <- c(3,3,3,4,3,1,2,3,1,3,4,4,2,3,3,1,3,3,4,3) # our data set
plot(x_1,type = 'o')
```


```{r}
mean_1 = mean(x_1) # calculate mean
var_1 = var(x_1)
mean_1
var_1
```

```{r}
x_2 <- c(1,2,3,4,5,1,2,3,1,2,4,5,2,3,1,1,2,3,5,6) # our data set
plot(x_2,type = 'b')
```
```{r}
(mean_2 = mean(x_2)) # calculate mean
(var_2 = var(x_2)) 
```

```{r}
plot(sort(x_2),type = 'p')
```


```{r}
summary(x_1)
```
```{r}
summary(x_2)
```

```{r}
plot(sort(x_2),type = 'p')
```

## Chebyshev Inequality 1

Suppose you collect call center wait times from a call center for 24 hours in a single day. Each hour, the mean and standard deviation of the call wait times will vary. Assume your call center is huge and you have 10,000 customer service representatives at any given hour.

Here are 24 data frames with 10,000 wait times per data frame. Each data frame was collected from a different hour on the same day.

```{r}
set.seed(123)
mu <- rpois(24, 5)  # true mean call center wait times
mu
list.df <- lapply(mu, function(x) rpois(10000, x))  # 10k call center workers per hour
str(list.df)
```

Given 24 data frames, you can calculate the sample mean and sample standard deviation for each hour.

```{r}
mean.each.hour <- sapply(list.df, mean)
sd.each.hour <- sapply(list.df, sd)
```

```{r}
mean.each.hour
sd.each.hour
```

You could also calculate the 95th percentile of each hour

```{r}
p95.each.hour <- sapply(list.df, function(x) quantile(x, probs = 0.95))
p95.each.hour
```

The question is: Are the observed 95th percentiles consistent with the theoretical relationship between the mean, standard deviation, percentiles? Assuming the sample mean and sample standard deviation are exactly equal to the population mean and population standard deviation.

We can use Chebyshev Inequality for this.

```{r}
Chebyshev.k <- function(rt_prob=0.05){k = sqrt(1 / rt_prob)}
k <- Chebyshev.k(0.05)
k
```


```{r}
install.packages('chebpol')
```


```{r}
Chebyshev.max <- function(means, stds, rt_probs){
  k <- Chebyshev.k(rt_probs)
  theoretical.max <- means + k*stds
  return(theoretical.max)
}
p95.theoretical <- Chebyshev.max(mean.each.hour, sd.each.hour, rt_probs=0.05)
p95.theoretical >= p95.each.hour
```

In our 24 data frames, each observed 95th percentile is less than or equal to the theoretical maximum given by Chebyshev Inequality.


##Chebyshev’s Theorem 3
Chebyshev’s theorem gives a lower bound for the area under a curve between two points that are on opposite sides of the mean and at the same distance from the mean.

For any number k greater than 1, at least 1−1/(k^2) of the data values lie within k standard deviations of the mean.

Let us use R to gain some intuition for Chebyshev’s theorem.


```{r}
k <- seq(1,4,by = 0.1)
auc <- 1-(1/k^2)
auc.percent <- round(auc*100)
cbind(k,auc.percent)
```

To put it in words: Let us pick a value for k: k=2. This means that at least 75% of the data values lie within 2 standard deviations of the mean.

Let us plot Chebyshev’s theorem with R:


```{r}
plot(k, 
     auc.percent, 
     col = 'blue', 
     pch = 19, 
     xlab = 'k', 
     ylab = 'percent', 
     main = 'Chebyshev\'s theorem' )
```

The theorem applies to both sample and population data. Note that Chebyshev’s theorem is applicable to a distribution of any shape. However, Chebyshev’s theorem can be used only for k>1. This is so because when k=1, the value of (1−1/k2) is zero, and when k<1, the value of (1−1/k2) is negative (Mann 2012).


